---
title: "Processing PLINK files"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(penalizedLMM)
```

The `penncath_lite` data is in the PLINK format (i.e., bed/bim/fam files), so I need to do some preprocessing first. The `process_plink` function uses functions from the [bigsnpr package](https://privefl.github.io/bigsnpr/reference/index.html) to do some quality control filtering steps. The result of the `process_plink` function is an object with 1) a matrix of genotypes and 2) a vector with the indices of any constant features (e.g., SNPs with no variation across the sample). 

Note: I am reading in the data using `plink_example()`, a function that serves as a wrapper to `base::system.file()`. This is simply because the exact filepath is different for the user than for the developer. If you save the PLINK files in a location other than the directory where you installed `penalizedLMM`, just adjust your choice of directory in the following function calls. 

```{r penncath_lite_process}
# preprocess PLINK files
penncath_lite <- process_plink(
  data_dir = plink_example(parent = TRUE),
  prefix = "penncath_lite",
  gz = TRUE, # NB: PLINK data that ships with package comes gzipped
  impute = TRUE, # mode imputation is default 
  outfile = "process_penncath")
```

```{r get_data}
# read in the data to global environment
pen <- get_data(path = paste0(plink_example(parent = TRUE), "/penncath_lite"))

str(pen) # Three objects here
```


In addition to the genotype data, I want to analyze a continuous phenotype stored in an external data file. I will load that here: 

```{r}
# NB: the 'quiet' option in process_plink()  will silence the printed messages
pen_cl <- read.csv(plink_example(path = 'penncath_clinical.csv'))

# for the sake of illustration, I use a simple mean imputation for the outcome 
pen_cl$hdl_impute <- ifelse(is.na(pen_cl$hdl),
                                       mean(pen_cl$hdl, na.rm = T),
                                       pen_cl$hdl)
```


Now that we have the data processed, we can fit a model. 

```{r penncath_lite_fit}
pen_fit <- plmm(X = pen$X,
                     y = pen_cl$hdl_impute)
# to see a progress bar and progress messages, set trace = TRUE
names(pen_fit)
```

We can summarize this fit at the 25th $\lambda$ value:  

```{r summary1}
summary(pen_fit, lambda = pen_fit$lambda[25])
```

And we can plot the coefficient paths; note that in these graphs, regions shaded in grey corresponds to values of $\lambda$ for which the model fit is nonconvex. See `ncvreg::plot.ncvreg()` for details.

```{r}
plot(pen_fit)
```

We can implement cross-validation for this model using the code below: 

```{r penncath_cv}
pen_cv <- cv.plmm(X = pen$X,
                       y = pen_cl$hdl_impute)
# to see progress bar, run the above with the trace = TRUE argument; this will
#   show a progress bar for each fold. 

print(summary(pen_cv, lambda = "min"))

plot(pen_cv)
```


More to come...
