---
title: "Choosing k -- an eigenproblem"
subtitle: "Close is only good in horseshoes..."
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(penalizedLMM)
library(corrplot)
```

## Context & motivation 
One of the novel features of the `penalizedLMM` package is in the way our methods address the relationships among samples (e.g., patients/participants) in the data. Unlike other mixed models packages like `lme4`, our package does not based relationships on a grouping factor. Rather, relationships are described as a matrix with cell values capturing the degree of relationship among each pair of samples. Let $\mathbf{X}$ represent the $n \times p$ design matrix of the data, and let $\tilde{\mathbf{X}}$ be the standardized $\mathbf{X}$ [1]. We call the $n \times n$ relationship matrix $\mathbf{K}$. By default, $\mathbf{K}$ is estimated from the data using the from data using $\mathbf{K} = \frac{1}{p} \tilde{\mathbf{X}}\tilde{\mathbf{X}}^\top$ [2]. 

The rotation that is used in `plmm_prep()` (an internal function in `plmm`) requires a singular value decomposition (SVD) of $\mathbf{K}$. This becomes an expensive calculation when $n$ is large. To improve the computational time of this calculation, we include an optional argument $k$ in our `plmm()` function -- this $k$ is the parameter to invoke *truncated* SVD. The truncated SVD of a matrix has been shown to be the optimal rank $k$ approximation to that matrix (Eckart-Young-Mirsky theorem) [3]. We use $\mathbf{A}_k$ to represent the rank $k$ approximation of $\mathbf{K}$, and we implement this approximation using truncated SVD using `RSpectra::svds()`.

While using truncated SVD can improve computational time by orders of magnitude, this approach raises the question of how to choose an appropriate value for $k$. This choice is not one that lends itself to a 'rule of thumb' -- how may $k$ it takes to approximate the SVD of $\mathbf{K}$ well depends on (1) the makeup of the specific data and (2) the user's definition of a 'close' approximation. 

To help users choose a suitable $k$ value to use, we offer the function `choose_k()`, which I will describe below with an extended example. 

## What this looks like in practice 

Consider again the `admix` genotype data with the simulated continuous outcome. Suppose I want to approximate $\mathbf{K}$. I could use the following: 

```{r}
approx1 <- choose_k(X = admix$X, returnKapprox = T, returnK = T)
```

Here we see that the starting value for $k$ is 19 -- by default, `choose_k()` makes the following 'decisions': 

  - The starting value for $k$ is set to be a tenth of the $n$ value. This is controlled by the `start` argument. 
  - Similarly, the default stepsize is also one tenth of the $n$ value. This is controlled by the `step` argument. 
  - The metric used to assess the quality of the approximation is the Frobenius norm. This is controlled by the `type` argument (see `Matrix::norm`). 
  - The $\epsilon$ value used to determine when $\delta \equiv \frac{||\mathbf{K} - \mathbf{A}_k||_{\text{F}}}{||\mathbf{K}||_{\text{F}}}$ is satisfactorily small enough is automatically set to 2. This is controlled by the `eps` argument. 
  
If the user wants to have a smaller $\epsilon$ and start at a higher value, this is another option: 

```{r}
approx2 <- choose_k(X = admix$X, start = 30, eps = 0.05, returnKapprox = T)
```

Notice that these apparent 'tweaks' in parameters have a notable impact on the choice of $k$ -- the balance of these parameters is both an art and a science. 

We can look at the $\mathbf{A}_k$ matrices produced by both approximations above, comparing to the true $\mathbf{K}$ (at far right):

```{r}
par(mfrow=c(1,3))
corrplot(cov2cor(approx1$K_approx),
         title = "Approx. 1",
         mar = c(0,0,2,0),
         tl.col = 'grey40',
         addgrid.col = NA, 
         tl.pos = "n")
corrplot(cov2cor(approx2$K_approx),
         title = "Approx. 2",
         mar = c(0,0,2,0),
         tl.col = 'grey40',
         addgrid.col = NA, 
         tl.pos = "n")
corrplot(cov2cor(approx1$K),
         title = "True K",
         mar = c(0,0,2,0),
         tl.col = 'grey40',
         addgrid.col = NA, 
         tl.pos = "n")
```

Now that we have some approximations, we can use `cv.plmm` to see the results: 

```{r}
fit1 <- cv.plmm(X = admix$X, y = admix$y, K = approx1$K_svd)
fit2 <- cv.plmm(X = admix$X, y = admix$y, K = approx2$K_svd)

par(mfrow=c(1,2))
plot(fit1)
plot(fit2)
```

```{r}
summary(fit1)
summary(fit2)
```


## References 

[1] Reisetter, A. C., & Breheny, P. (2021). Penalized linear mixed models for structured genetic data. Genetic Epidemiology, 45(5), 427-444.

[2] CITATION NEEDED for standardization 

[3] Eckart, C., & Young, G. (1936). The approximation of one matrix by another of lower rank. Psychometrika, 1(3), 211-218.
