---
title: "If your data is in a delimited file"
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(plmmr)
```

If you have data stored as a delimited file (e.g., a `.txt` file), this is the place for you to begin. To analyze such data, there is a 3-step procedure: (1) process the data, (2) create a design, and (3) fit a model. 

## Process the data 

```{r}
 # I will create the processed data files in a temporary directory; 
#   fill in the `rds_dir` argument with the directory of your choice
temp_dir <- tempdir()

colon_dat <- process_delim(data_file = "colon2.txt",
  data_dir = find_example_data(parent = TRUE), 
  rds_dir = temp_dir,
  rds_prefix = "processed_colon2",
  sep = "\t",
  overwrite = TRUE,
  header = TRUE)

# look at what is created 
colon <- readRDS(colon_dat)
```

The output messages indicate that the data has been processed. This call created 2 files, one `.rds` file and a corresponding `.bk` file. The `.bk` file is a special type of binary file that can be used to store large data sets. The `.rds` file contains a pointer to the `.bk` file, along with other meta-data. 

Note that what is returned by `process_delim()` is a character string with a filepath: `r str(colon_dat)`. 

## Create a design 

Creating a design ensures that data are in a uniform format prior to analysis. For delimited files, there are two main processes happening in `create_design()`: (1) standardization of the columns and (2) the construction of the penalty factor vector. Standardization of the columns ensures that all features are evaluated in the model on a uniform scale; this is done by transforming each column of the design matrix to have a mean of 0 and a variance of 1. The penalty factor vector is an indicator vector in which a 0 represents a feature that will always be in the model -- such a feature is *unpenalized*. To specify columns that you want to be unpenalized, use the 'unpen' argument. Below in our example, I am choosing to make 'sex' an unpenalized covariate.  

A side note on unpenalized covariates: for delimited file data, all features that you want to include in the model -- both the penalized and unpenalized features -- must be included in your delimited file. This differs from how PLINK file data are analyzed; look at the `create_design()` documentation details for examples. 

```{r}
# prepare outcome data
colon_outcome <- read.delim(find_example_data(path = "colon2_outcome.txt"))

# create a design
colon_design <- create_design(data_file = colon_dat,
                              rds_dir = temp_dir,
                              new_file = "std_colon2",
                              add_outcome = colon_outcome,
                              outcome_id = "ID",
                              outcome_col = "y",
                              unpen = "sex", # this will keep 'sex' in the final model
                              logfile = "colon_design")
```

As with `process_delim()`, the `create_design()` function returns a filepath: `r str(colon_design)`. The output messages document the steps in the create design procedure, and these messages are saved to the text file `colon_design.log` in the `rds_dir` folder. 

For didactic purposes, we can look at the design: 

```{r}
# look at the results
colon_rds <- readRDS(colon_design)
str(colon_rds)
```


## Fit a model 

We fit a model using our design as follows: 

```{r}
colon_fit <- plmm(design = colon_design, return_fit = TRUE, trace = TRUE)
```

Notice the messages that are printed out -- this documentation may be optionally saved to another `.log` file using the `logfile` argument. 

We can examine the results at a specific $\lambda$ value: 

```{r}
summary(colon_fit, idx = 50)
```

We may also plot of the paths of the estimated coefficients: 

```{r}
plot(colon_fit)
```

## Prediction for filebacked data

This example shows an experimental option, wherein we are working to add a prediction method for filebacked outside of cross-validation. 

```{r}
# linear predictor 
yhat_lp <- predict(object = colon_fit,
        newX = attach.big.matrix(colon$X),
        type = "lp")

# best linear unbiased predictor 
yhat_blup <- predict(object = colon_fit,
        newX = attach.big.matrix(colon$X),
        type = "blup")

# look at mean squared prediction error 
mspe_lp <- apply(yhat_lp, 2, function(c){crossprod(colon_outcome$y - c)/length(c)})
mspe_blup <- apply(yhat_blup, 2, function(c){crossprod(colon_outcome$y - c)/length(c)})
min(mspe_lp)
min(mspe_blup)

# Pick up here: need to double-check the BLUP method....
```

