---
title: "Types of pointers for filebacked objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Types of pointers for filebacked objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(plmmr) # also loads bigmemory package
library(bigstatsr)
```

The technique of 'filebacking' allows us to 'point' to data on disk, rather than reading data into the random access memory (RAM).

`plmmr` uses two different kinds of 'pointers' to access data which are too large to read into RAM. The table below contrasts the two pointers (in a high-level view):

| bigstatsr::FBM                                                                                                                                                                | bigmemory::big.matrix                                                                                                                    |
|------------------------------------|------------------------------------|
| allows non-contiguous subsetting (essential for cross-validation)                                                                                                             | only allows contiguous subsets                                                                                                           |
| subset method creates a copy (takes up space)                                                                                                                                 | subset method does *not* create a copy (saves space)                                                                                     |
| easily converts to `big.matrix`                                                                                                                                               | can be set up to use the same backingfile as an `FBM`, but conversion can be 'buggy'                                                     |
| cannot be multiplied directly with a `Matrix::dgCMatrix` object                                                                                                               | can be multiplied with a `Matrix::dgCMatrix` object – essential for making predictions                                                   |
| not designed to work with `bigalgebra` directly; the `bigalgebra` arithmetic methods don't work with `FBM` objects                                                            | seamless interaction with companion package `bigalgebra`, which allows for efficient computation arithmetic tasks like rotating the data |
| seamless interaction with companion package `bigsnpr` – lots of infrastructure specifically designed for genetic/genomic data (including methods to read/process PLINK files) | not designed to work directly with PLINK files                                                                                           |

: FBM v. big.matrix pointers

The table illustrates that these two kinds of pointers each have pros and cons – and \`plmmr\` draws on the strengths of both in order to improve computational efficiency and offer infrastructure for genetic data.

Below is a short demo that illustrates how these two different kinds of pointers can work together. **Note**: for the sake of clarity, this demo ignores/brushes over some standardization details; while these details don't affect the pointer type(s), these details are **very** important for model fitting. Please, don't blindly follow this demo as a guide for analysis.

```{r}
# an FBM with genetic data
admix$fbm_X <- bigstatsr::as_FBM(admix$X)
dim(admix$fbm_X)
class(admix$fbm_X)
admix$fbm_X$backingfile

# now, to get ready for other plmm() computation, I switch to a big.matrix:
admix$bm_X <- fbm2bm(admix$fbm_X)
# notice: the backing file is the same. Both pointers are pointing to the same data:
paste0(bigmemory::dir.name(admix$bm_X), bigmemory::file.name(admix$bm_X))

# pretend we need to do cross-validation. Back to FBM for non-contiguous subsetting:
fold <- sample(1:nrow(admix$bm_X) %% 10) # choose 10 for 10 folds
fold[fold==0] <- 10

# hold out 1st fold: 
training_data <- bigstatsr::big_copy(admix$bm_X,
                                     ind.row = which(fold!=1))
dim(training_data)
class(training_data) # big_copy() returns an FBM
training_data$backingfile # subset data is copied with a new backing file

# now within each fold, we need to rotate:
admix$K <- relatedness_mat(admix$X[which(fold!=1),])
eig_K <- eigen(admix$K)
eta <- 0.5 
w <- (eta * eig_K$values + (1 - eta))^(-1/2)
wUt <- sweep(x = t(eig_K$vectors), MARGIN = 1, STATS = w, FUN = "*")
training_data <- fbm2bm(training_data)
rot_X <- wUt %*% training_data # here, %*% operator is from bigalgebra
dim(rot_X)
```
